\chapter{ML Functions in the wellorder Library}
\label{entries}
\input{entries-intro}
\DOC{set\_wo\_map}

\TYPE {\small\verb%set_wo_map : (void -> void)%}\egroup

\SYNOPSIS
Sets interface map for constants in the ``wellorder'' library.

\DESCRIBE
To increase flexibility, the constants used in the wellorder library are
declared with a {\small\verb%wo_%} prefix. The {\small\verb%set_wo_map%} function sets up an interface
map to eliminate them from view. Any existing interface map is preserved
alongside. The constants which have their prefixes hidden are: {\small\verb%subset%},
{\small\verb%Union%}, {\small\verb%fl%}, {\small\verb%poset%}, {\small\verb%chain%}, {\small\verb%woset%} and {\small\verb%inseg%}.

\FAILURE
Fails if an entry in the new interface map clashes with an existing one.

\COMMENTS
The interface map is not set up by default when the library is loaded.

\SEEALSO
interface_map, set_interface_map, unset_wo_map.

\ENDDOC
\DOC{unset\_wo\_map}

\TYPE {\small\verb%unset_wo_map : (void -> void)%}\egroup

\SYNOPSIS
Removes interface map for constants in the ``wellorder'' library.

\DESCRIBE
To increase flexibility, the constants used in the wellorder library are
declared with a {\small\verb%wo_%} prefix. The {\small\verb%unset_wo_map%} function cancels the effect of
{\small\verb%set_wo_map%}, which sets up an interface map to eliminate them from view. Any
existing interface map is preserved when the wellorder-related entries are
eliminated. The constants which have their prefixes hidden are: {\small\verb%subset%},
{\small\verb%Union%}, {\small\verb%fl%}, {\small\verb%poset%}, {\small\verb%chain%}, {\small\verb%woset%} and {\small\verb%inseg%}.

\FAILURE
Fails only if some entries in the interface map matching the above have been
set by other means, so if subtracted, a constant would be hidden.

\COMMENTS
The interface map is not set up by default when the library is loaded.

\SEEALSO
interface_map, set_interface_map, set_wo_map.

\ENDDOC
